---
title: "Data Analysis Assignment 2"
subtitle: "COMM 7370 | Spring 2023 | Due: see syllabus"
format:
  pdf:
    geometry: 
      - top=.75in
      - left=.8in
      - right=.8in
    urlcolor: blue
  html:
    code-fold: true
    toc: true
    theme: sandstone
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(kableExtra)
library(summarytools)
library(rstatix)
library(formatR)

# Read HELP data into R
hdata <- read_csv("https://raw.githubusercontent.com/sarakyeo/sarakyeo.github.io/master/COMM-3710/LA/HELP.csv")
```

In this assignment, you will learn how to:

-   Examine frequency distributions using the `freq()` function.
-   Select a subset of data based on criteria using the `filter()` function.
-   Recode variables using the `mutate()` and `case_when()` functions.
-   Create new variables from existing variables in a data frame.

::: callout-important
## Remember to\...

1.  Set up your R script as you did in your previous assignment.

2.  Include pseudocode in your R script.
:::

------------------------------------------------------------------------

# Codebook

| Variable     | Description                                                                                                       |
|------------------------|------------------------------------------------|
| `age`        | R age at baseline                                                                                                 |
| `anysub`     | Use of any substance post-detox (no, yes)                                                                         |
| `cesd`       | Center for Epidemiologic Studies Depression measure at baseline (higher scores indicate more depressive symptoms) |
| `d1`         | Lifetime number of hospitalizations for medical problems (measured at baseline)                                   |
| `daysanysub` | Time (in days) to first use of any substance post-detox                                                           |
| `dayslink`   | Time (in days) to linkage to primary care                                                                         |
| `drugrisk`   | Risk Assessment Battery drug risk scale at baseline                                                               |
| `e2b`        | Number of times in the past 6 months R entered a detox program (measured at baseline)                             |
| `female`     | Biological sex (female coded high)                                                                                |
| `sex`        | Biological sex (male, female)                                                                                     |
| `g1b`        | Experienced serious thoughts of suicide in last 30 days (measured at baseline)                                    |
| `homeless`   | Housing status (housed, homeless)                                                                                 |
| `i1`         | Average number of drinks consumed per day in past 30 days (measured at baseline)                                  |
| `i2`         | Maximum number of drinks consumed per day in past 30 days (measured at baseline)                                  |
| `id`         | R identifier                                                                                                      |
| `indtot`     | Inventory of Drug Use Consequences (InDUC) total score (measured at baseline)                                     |
| `linkstatus` | Post-detox linkage to primary care (yes coded high)                                                               |
| `link`       | Post-detox linkage to primary care (no, yes)                                                                      |
| `mcs`        | SF-36 Mental Component Score (measured at baseline; lower scores indicate worse status)                           |
| `pcs`        | SF-36 Physical Component Score (measured at baseline; lower scores indicate worse status)                         |
| `pss_fr`     | Perceived social support by friends (measured at baseline; higher scores indicate more support)                   |
| `racegrp`    | Race/Ethnicity (black, hispanic, white, other)                                                                    |
| `satreat`    | Any BSAS substance abuse treatment at baseline (no, yes)                                                          |
| `sexrisk`    | Risk Assessment battery                                                                                           |
| `substance`  | Primary substance of abuse (alcohol, cocaine, heroin)                                                             |
| `treat`      | Randomized to HELP clinic (no, yes)                                                                               |

: : Codebook for the HELP dataset. {#tbl-codebook}

# Instructions

1)  Load the following packages in R and read in the [HELP data](https://raw.githubusercontent.com/sarakyeo/sarakyeo.github.io/master/COMM-3710/LA/HELP.csv) using the `read_csv()` function.[^1]

    -   `tidyverse`

    -   `summarytools`

    -   `rstatix`

        ```{r}
        #| include: false

        hdata <- read.csv("https://raw.githubusercontent.com/sarakyeo/sarakyeo.github.io/master/COMM-3710/LA/HELP.csv")

        ```

2)  Often, the first step in data analysis is to examine variables of interest. To do so, we use frequency distributions. To create a frequency table, we can use the `freq()` function in the `summarytools` package. For example, if I want to examine the frequency distribution of variable, `var1`, in the data frame, `df1`, I would use the following code:

    ```{r}
    #| eval: false
    #| echo: true

    df1 %>% 
            freq(var1) # Examine freq dist of var1 in df1 data frame

    ```

    Now, examine the frequency distributions of the variables `sex` and `d1` from the HELP data.

    a)  How many patients in the study are female?

    b)  How many patients in the study have never been hospitalized for medical problems?

    c)  What percentage of patients in the study have been hospitalized fewer than 5 times (i.e., 4 or fewer)?

    ::: callout-note
    ## XQuartz (for Mac users)

    If you are having trouble with the `summarytools` package, you may see the following error message:

    ```{r}
    #| eval: false
    #| include: true

    Error: package or namespace load failed for ‘summarytools’:
    .onLoad failed in loadNamespace() for 'tcltk', details:
    call: fun(libname, pkgname)
    error: X11 library is missing: install XQuartz from xquartz.org

    ```

    To resolve this error, simply read the warning and follow the instructions (i.e., install XQuartz from the source provided).

    If you continue to experience issues with `summarytools`, try reinstalling XQuartz. If you continue to encounter problems, install `summarytools` directly from Github in R using the following code:

    ```{r}
    #| eval: false
    #| include: true

    install.packages("devtools")
    devtools::install_github("dcomtois/summarytools",
                             ref = "no-x11-check")

    ```
    :::

3)  Next, we will learn to subset the data to include only respondents who meet certain criteria. To do so, we use the `filter()` function. If, for example, I want to examine the ages of respondents in the HELP data who are female and have had fewer than 10 hospitalizations, I would use the following code:

    ```{r}
    #| eval: false
    #| include: true

    hdata %>%
            filter(sex == "female" & d1 < 10) %>%
            freq(age)

    ```

    Your turn: Subset the data to include respondents whose primary substance of abuse is cocaine and who are at least 40 years old. Then, examine a frequency distribution of `age` among this subset.

    a)  How many patients are included in this subset?

    b)  What is the mean age in this subset? **Hint:** Use the `descr()` function to get descriptive statistics.

4)  Often, we want to recategorize continuous variables as categorical ones. We call this process [**recoding variables**]{.underline}. For example, I want to categorize respondents by their SF-36 Physical Component Score, `pcs`. I want to create a new variable, `dpcs`, with two categories, where respondents with low `pcs` are those who scored 40 or less on this measure. To do so, I would use the `mutate()` and `case_when` functions:

    ```{r}
    #| eval: false
    #| include: true

    hdata <- hdata %>%
            mutate(dpcs = case_when(pcs <= 40 ~ "low",
                                    pcs > 40 ~ "high"))

    # To check that I have created the new variable, dpcs, correctly, I use freq()
    hdata %>% 
            freq(dpcs)

    ```

    Your turn: Recode respondents with depressions scores of 30 or lower into a low category and those who have scores higher than 30 into a high category. In other words, recode `cesd` into a new variable, `dcesd` (or call it whatever you would like), with only two categories, `low` ($\le$ 30) and `high` (\> 30).

    a)  How many respondents are in each category?

    b)  What is the mean age of respondents in each of these categories? **Hint:** You can use the `filter()` and `descr()` functions to answer this question.

5)  Patients with a mental component score (`mcs`) less than 20 are thought to be at extreme risk of returning to the detoxification unit within the next 12 months. Make a new variable called `ExtremeMCS` and code it as `1` if a patient is at risk based on his/her `mcs` score and `0` otherwise. Then, answer the following questions:

    a)  How many patients are at risk of returning to the detoxification unit in the next 12 months?

    b)  What percentage of patients are at low risk of returning to the detox unit within the next 12 months?

6)  Create two new variables, `SuicidalThought` and `HomelessStatus` based on `g1b` and `homeless`, respectively. If a patient has not experienced thoughts of suicide in the last 30 days, code them as `0` for `SuicidalThought`. If a patient is housed, code them as `0` for `HomelessStatus`. We will use these new variables, along with `ExtremeMCS` to create a scale of risk factors for each patient.

    Suppose `ExtremeMCS`, `SuicidalThought`, and `HomelessStatus` are considered risk factors. Construct a new variable called `RiskTotal` that quantifies the number of risk factors for each patient (i.e., make it a sum of these 3 variables).

    a)  What percentage of patients in the study have fewer than 3 risk factors?

    ::: callout-tip
    To create a new variable, `newVar`, that is the sum of existing variables, `var1` and `var2`, use the `mutate()` function.

    ```{r}
    #| eval: false
    #| include: true

    dataframe <- dataframe %>% # write the new variable back into the dataframe
            rowwise() %>% 
            mutate(newVar = var1 + var2) # sum to get new variable

    # Note the inclusion of the function rowwise() here with nothing in the ().
    # We include this to tell R to add the variables row-by-row.

    dataframe %>% 
            freq(newVar) # check that newVar is created correctly

    ```

    

    :::

[^1]: If you need a reminder of how to do this, check Data Analysis Assignment 1.

# Submission

Submit your R script (which should have a `.R` extention) to Canvas. Your R script should:

1)  Include code to install and load the packages.
2)  Contain comments and/or pseudocode.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Completely shut down RStudio or restart your R session.
-   Reopen RStudio and your `.R` script.
-   Run the entire script by clicking the "Run" button in the top right of the R script.

::: callout-important
**These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for *all* of your data analysis assignments.**
:::
