---
title: "Data Analysis Assignment 1"
subtitle: "COMM 7370 | Spring 2023 | Due: see syllabus"
format:
  pdf:
    geometry: 
      - top=.75in
      - left=.8in
      - right=.8in
    urlcolor: blue
  html:
    code-fold: true
    toc: true
    theme: sandstone
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(kableExtra)
library(summarytools)
library(rstatix)
library(formatR)

# Read HELP data into R
hdata <- read_csv("https://raw.githubusercontent.com/sarakyeo/sarakyeo.github.io/master/COMM-3710/LA/HELP.csv")
```

In this assignment, you will learn how to:

-   Prepare your R script for your assignment.
-   Create foldable/collapsible headers in your R script.
-   Install and load packages in R.
-   Open a help page in R.
-   Manually enter data into R and read/load data into R from a .csv file.
-   Identify variables and their types in a tibble.
-   Use the pipe operator,`|>` or `%>%`.

------------------------------------------------------------------------

# Instructions

1)  Read [Ch. 3 of COMM 3710: Getting Started with R](https://sarakyeo.github.io/COMM-3710-bookdown/docs/setup.html). Then, create a new R script. Prepare your R script. In the first four (4) lines of your script, include the following, each on a new line, as a comment:

    -   Full name

    -   Student ID (uXXXXXXX)

    -   Date (MM/DD/YY)

    -   Assignment name

    ::: callout-tip
    **Comments** (which are preceded by `#` in your R script) are not run or executed by R. R will only run/execute **commands a.k.a. functions** (not preceded by `#`).
    :::

```{=html}
<!-- -->
```
2)  Make a section of foldable code titled "Installing and loading packages." Under this heading, **write pseudocode** in preparation to install and load the following packages:

    -   `tidyverse`

    -   `summarytools`

    -   `rstatix`

    ::: callout-tip
    **Pseudocode** is the plain English version of your R code that described the steps you are taking in R. It should be written as comments in your R script and should be logical.
    :::

    ::: callout-warning
    While you are welcome to work with another student on your assignments, your pseudocode **should be in your own words**. Identical submissions will be flagged as plagiarism, which will be dealt with in accordance with the course policy.
    :::

    Then, write commands to **install** the above packages and **load** them into your current session of R.

    ::: callout-tip
    [Sections 2.3.5 and 2.3.6](https://sarakyeo.github.io/COMM-3710-bookdown/docs/start.html#packages) in COMM 3710: Getting Started with R should help.
    :::

3)  Save your R script as `LA-3_FirstName-LastName.R`.

4)  In the next step, we will enter data into R manually using the `c()` function, which is short for `combine()`. Before we do so, let's learn about the function we will be using. Create a new foldable section in your code for this--give it an appropriate heading. Write pseudocode to pull up the help documentation for the `c()` function (e.g., "exploring the `c()` function before using it"). Then, write and run the R command (`?c`) that opens the documentation in your R script. Use the results to describe the `c()` function from the documentation as a comment under your `?c` command. You can copy and paste the text under "Description" from the documentation, but I encourage you to add your own description so that you understand what the function is doing.

    ::: callout-tip
    Typing `?` before any function in R will open the documentation for that function. I recommend that you use this any time you are unsure what a R function does. For example, if I wanted to learn more about the `library()` function, I would type and run the command, `?library`.
    :::

5)  Time to enter data! Write pseudocode to enter the data in the table below (Table @tab-data) as four (4) objects called `Aname`, `Ateam`, `ppg`, and `pts`, respectively. Writing pseudocode as comments before actual code should become a habit as it will help you understand how to write appropriate code for data analysis. After writing code in your R script, remember to **run** it.

    ```{r}
    #| label: tab-data
    #| tab-cap: "Data from 2021 of top 5 women's college basketball players by points per game."
    #| echo: false
    #| warning: false

    Aname <- c("C. Clark", "A. Hayes", "C. Hooks", "K. Bell", "A. Joens")
    Ateam <- c("Iowa", "Middle Tenn", "Ohio", "FGCU", "Iowa State")
    ppg <- c(26.6, 26.5, 25.1, 24.3, 24.2)
    pts <- c(799, 663, 628, 632, 678)
    df <- tibble(Aname, Ateam, ppg, pts)

    df %>% 
      kable() 
    ```

    ::: callout-note
    ## Data Types

    There are two types of data in Table @tab-data, numeric and character. To enter these data into objects in R as the appropriate type of data, review [Section 2.4](https://sarakyeo.github.io/COMM-3710-bookdown/docs/start.html#learning-by-doing) in the book.
    :::

    Check that you have created the objects correctly by calling each object. Calling an object in R allows you to see it in the Console. Write pseudocode for this, followed by the R command to call each object (`Aname`, `Ateam`, `ppg`, `pts`).

    Then, combine the objects into a single data frame that has a name of your choosing and look at the dataframe. Use the `tibble()` function to do this. Again, pseudocode and R code should be included in your script. The data in the tibble should match the information in Table @tab-data.

    ::: callout-tip
    To call an object, type the name of the object in the Console and press Enter. Or type the name of the object in your R script, then highlight and run it.
    :::

    The pseudocode and R code below is generic. This means that you will have to replace the placeholders (e.g., `obj1`, `obj2`, `df`) with names of the objects in your R environment.

    ```{r}
    #| eval: false

    # Calling obj1 to check that I have created it correctly
    obj1

    # Put obj1, obj2, etc. into a tibble (tidyverse-style data frame) 
    df <- tibble(obj1, obj2, obj3, ...)
    ```

6)  Now, we will examine the type of data stored in each column of our data frame. To do this, we will use the `class()` function. First, write pseudocode and R code to examine the `class` function. Copy and paste the description of this function as a comment in your R script. Then, write pseudocode and R code to check the class of each column in your data frame. To reference a column in a data frame, we use one type of extractor operator, `$`. Let's say I have a data frame called `cats` and I wanted to examine the data type of a column within the `cats` data frame called `cuteness`. To do so, I will use the class function and an extractor operator: `class(cats$cuteness`.

    Run the code you just wrote and note the class of each column of your data frame in a comment. If necessary, review the [brief section in the book about data types](https://sarakyeo.github.io/COMM-3710-bookdown/docs/start.html#data-types).

For this last step, we will use a Tutorial that is built into an existing R package.

Install and load the `learnr` package in your current session of R. In the **Tutorial** panel (usually top right with the **Environment** panel), complete the **Data Basics** tutorial.

Include the answers to the questions below in your R script as comments. Clearly label your answers with the question number (e.g., 10a, 10b). You will answer these questions as part of the Data Basics Tutorial. List the answers from your tutorial as comments in your R script.

a)  What does the `drv` variable of `mpg` describe? Read the help for `?mpg` to find out.
b)  How many rows are in the data frame named `cars`?
c)  How many columns are in the data frame named `cars`?
d)  Which types of variables does `flights` contain? List all the types that you checked in the tutorial.

------------------------------------------------------------------------

# Submission

Submit your R script (which should have a `.R` extention) to Canvas. Your R script should:

1)  Include code to install and load the packages.
2)  Contain comments and/or pseudocode.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Completely shut down RStudio or restart your R session.
-   Reopen RStudio and your `LA-3_FirstName-LastName.R` script.
-   Run the entire script by clicking the "Run" button in the top right of the R script.

::: callout-important
**These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for *all* of your data analysis assignments.**
:::
