---
title: "Practice Data Wrangling"
format:
   pdf:
      documentclass: article
      geometry:
         - top=20mm
         - left=20mm
         - right=20mm
         - heightrounded
      toc: false
      colorlinks: true
urlcolor: blue
---

# Instructions

::: callout-tip
It is often helpful to read the assignment in its entirety before starting to work on it.
:::

1)  Set up your R script for this assignment (name, uNID, date, assignment number). Install (if necessary) and load the packages below.

-   `tidyverse`
-   `summarytools`
-   `rstatix`

```{r}
#| label: ans1
#| include: false

library(tidyverse)
library(summarytools)
library(rstatix)

```

2)  Read `broadway.csv` into R like you did in LA-4 using the `read_csv()` function.

```{r}
#| label: ans2
#| include: false

bway <- read_csv("broadway.csv")

```

3)  Use the pipe operator (`|>`), and the `filter()` and `summarise()` functions to answer the following questions:
    a)  What is the minimum number of people who attended *Miss Saigon*?
    b)  What is the maximum number of people who attended *Miss Saigon*?
    b)  What is the mean number of people who attended *Miss Saigon*?

4)  How many musicals and plays were there at the Stephen Sondheim theatre in 2011?

------------------------------------------------------------------------

# Submission

Submit your R script (named `AssignmentName_FirstName-LastName.R`) to Canvas.

Your R script should:

1)  Include commands and functions that are necessary to address all the questions in the assignment.
2)  Contain comments that answer the questions in the assignment.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Navigate back to Your Workspace on Posit Cloud.
-   Reopen your project.
-   Run the entire script line-by-line without editing it to ensure there are no errors.

::: callout-important
These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for all of your individual lab assignments.
:::
