---
title: "LA-5: Examining and Managing Data (15 points)"
format:
   pdf:
      documentclass: article
      geometry:
         - top=20mm
         - left=20mm
         - right=20mm
         - heightrounded
      toc: false
      colorlinks: true
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(summarytools)
library(rstatix)
library(kableExtra)
library(formatR)

# Read data into R
wind <- read_csv("wind-turbines.csv")

```

# Learning Outcomes

In this assignment, you will learn how to:

-   Run and examine frequency distributions using the `freq()` function in the `summarytools` package.
-   View descriptive statistics using the `descr()` function in the `summarytools` package.
-   Recoding variables using the `case_when()` function.
-   Create new variables from existing variables in a data frame using the `mutate()` function.

::: callout-tip
Read all the instructions carefully before starting the assignment.
:::

------------------------------------------------------------------------

# Instructions

1)  Install (`install.packages()`) and load (`library()`) the packages below. Then download `wind-turbines.csv` dataset from Canvas and read it in to R. You will also want to download the codebook for the data from Canvas.

    -   `tidyverse`
    -   `summarytools`
    -   `rstatix`

2)  The dataset includes data about wind turbines in the US. Use the `freq()` function to examine the frequency distributions of the variables `Site.State` and `Year`. Use the frequency distributions to answer the following questions.

    a)  How many wind turbines are located in Utah?
    b)  How many wind turbines became operational in 2021?
    c)  What percentage of turbines became operational between 1983 and 1999 (include these years in the calculation)?

```{r}
#| label: ans2
#| include: false
#| echo: false
#| eval: false

wind |> freq(Site.State)

# 2a: 205 turbines are in Utah.

wind |> freq(Year)

# 2b: 1027 turbines became operational in 2021.

wind |> 
        filter(Year >= 1983 & Year <= 1999) |> 
        freq(Year)

# 2c: 3.32%

```

3)  Now, subset the data to answer the following question: How many turbines in Utah have been operational since 2010?

```{r}
#| label: ans3
#| include: false
#| echo: false
#| eval: false

wind |> 
        filter(Year >= 2010 & Site.State == "UT") |> 
        freq(Site.State)

#3: 96

```

4)  Examine the frequency distribution of `Turbine.Capacity` based on the subset in #3.

    a)  How many turbines in this subset have a capacity of 2,300 kW?
    b)  What percentage of turbines in this subset have a capacity of 1,500 kW?

```{r}
#| label: ans4
#| include: false
#| echo: false
#| eval: false

wind |> 
        filter(Year >= 2010 & Site.State == "UT") |> 
        freq(Turbine.Capacity)

#4a: 27
#4b: 70.8%

```

5)  Often, we want to look at descriptive statistics of a quantitative variable. We can do this using the `descr()` function that is in the package `summarytools`. Examine the descriptive statistics for `Turbine.Total_Height` and answer the questions below. Include the units of measurement in your answers.

    a)  What is the mean?
    b)  What is the median?
    c)  What is the maximum?

```{r}
#| label: ans5
#| include: false
#| echo: false
#| eval: false

wind |> 
        descr(Turbine.Total_Height)

#5a: 127.63 m
#5b: 130.10 m
#5c: 199.60 m

```

6)  Next, we will learn how to use the function, `mutate()`. We use this function to create new variables (or "mutate" them) from existing ones. Let's create a new variable called `Turbine_Space` that is the mean of two existing variables, `Turbine.Hub_Height` and `Turbine.Rotor_Diameter`. Then, examine the descriptive statistics of the new variable, `Turbine_Space`, and answer the following questions:

    a)  What is the mean of `Turbine_Space`?
    b)  What is the minimum value of `Turbine_Space`?
    c)  What is the maximum value of `Turbine_Space`?

7)  Sometimes, we may want to make continuous variables into categorical variables. For example, let's say we are interested in the number of turbines in a project, `Project.Number_Turbines`, but we want to categorize them into low vs. high numbers of turbines in projects.\
    \
    We refer to this process as **recoding variables**. For this exercise, let's recode projects with 100 turbines or less into the `low` category and those that have more than 100 turbines into the `high` category.\
    \
    Use the `mutate()` and `case_when()` functions to recode `Project.Number_Turbines` into a new variable, `Project.Number`, with only two categories, low ($\le$ 100) and high ($>$ 100). Then, answer the following questions.

    a)  How many projects are in the **low** category?
    b)  How many projects are in the **high** category?

8)  **BONUS:** In the previous step, you created a new variable called `Project.Number`. What is the mean `Project.Capacity` among projects in the (a) low and (b) high `Project.Number` categories?

------------------------------------------------------------------------

# Submission

Submit your R script (named `LA-#_FirstName-LastName.R`) to Canvas.

Your R script should:

1)  Include commands and functions that are necessary to address all the questions in the assignment.
2)  Contain comments that answer the questions in the assignment.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Navigate back to Your Workspace on Posit Cloud.
-   Reopen your project.
-   Run the entire script line-by-line without editing it to ensure there are no errors.

::: callout-important
These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for all of your individual lab assignments.
:::
