---
title: "LA-6: Data Visualization (15 points)"
format:
   pdf:
      documentclass: article
      geometry:
         - top=20mm
         - left=20mm
         - right=20mm
         - heightrounded
      toc: false
      colorlinks: true
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(summarytools)
library(rstatix)
library(kableExtra)
library(formatR)

# Read data into R
util <- read_csv(here::here("COMM-3710", "LA","Utilities.csv"))

```

# Learning Outcomes

In this assignment, you will:

- Learn how to use `ggplot2`, which is a package included in the `tidyverse` suite of packages, to create visualizations using data.
- Practice using functions from LA-5, including `mutate()`, `freq()`, and `descr()`.

::: callout-tip
Read all the instructions carefully before starting the assignment.
:::

------------------------------------------------------------------------

# Instructions

1) Load the packages below and download the `Utilities.csv` data file from Canvas. Read the dataset into R. Use the codebook on Canvas to familiarize yourself with the data in this file.
   
   * `tidyverse`
   * `summarytools`

2) Read and follow along with Chapter 5 of [this online book](https://sarakyeo.github.io/COMM-3710-bookdown/docs/ggplot2.html). Use the data provided in the chapter to follow the steps in Ch. 5.

:::callout-tip
Do not skip this step. It will help you understand `ggplot2`.
:::

3) Make a histogram to display the distribution of customers' total monthly bill.

   a) Use the histogram you created and describe the distribution. Is the distribution skewed or symmetric? If skewed, is the distribution skewed positively or negatively?
   b) At what cost do most of the total monthly bills fall? Provide a range between which the total monthly bills fall.
   
```{r}
#| label: ans3
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = totalbill)) +
        geom_histogram()

#3a: Distribution is skewed positively.
#3b: Most of the total monthly bills fall around $100 to $150.

```

4) Make a scatterplot to display gas bill by month. Use your graph to answer the following questions.

   a) Which month do you think has the highest average gas bill?
   b) Use R to calculate the average gas bill for the months of January and December. Which is higher?
   
```{r}
#| label: ans4
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = month, y = gasbill)) +
        geom_point()

#4a: December

util |> 
        filter(month == 1) |> 
        descr(gasbill) # M = 180.38

util |> 
        filter(month == 12) |> 
        descr(gasbill) # M = 179.28

#4b: Average gas bills are higher in January than December.

```

5) Make a scatterplot to display electric bill by month. Use your graph to answer the following questions.

   a) Which month do you think has the highest average electric bill?
   b) Calculate the average electric bill for the months of September and December. Which is higher?
   
```{r}
#| label: ans5
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = month, y = elecbill)) +
        geom_point()

#5a: December

util |> 
        filter(month == 9) |> 
        descr(elecbill) # M = 99.16

util |> 
        filter(month == 12) |> 
        descr(elecbill) # M = 87.21

#5b: Average electricity bills are higher in September than December.

```

6) Make a scatter plot to display the relationship between gas usage and gas bill. Place gas usage on the x-axis and  monthly gas bill on the y-axis. Does there appear to be a relationship between gas usage and gas bill? If so, describe the relationship (e.g., does a relationship exist? Is it linear? Is there a positive or negative relationship between the variables?).

```{r}
#| label: ans6
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = ccf, y = gasbill)) +
        geom_point()

# Yes, there appears to be a relationship between gas bill and usage. There 
# appears to be a positive linear relationship between these two variables.

```

7) Make a scatterplot to show the relationship between electricity usage and electric bill. Does there appear to be a relationship between electricity usage and electric bill? If so, describe the relationship.

```{r}
#| label: ans7
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = kwh, y = elecbill)) +
        geom_point()

# Yes, there appears to be a relationship between electricity bill and usage.
# There appears to be a positive linear relationship between these two variables.

```

8) Create a new categorical variable called `season`. Have the variable equal `winter` if the bill was from December, January, or February. Have the variable equal `spring` if the bill was from March, April, or May. Have the variable equal `summer` if the bill was from June, July, or August. Have the variable equal `fall` if the bill was from September, October, or November.\
\
Then, examine a frequency table of `season`.

   a) How many bills in the data set are from fall?
   b) How many bills in the data set are from summer?
   
```{r}
#| label: ans8
#| include: false
#| eval: false

util <- util |> 
        mutate(season = case_when(
                month == 12 | month == 1 | month == 2 ~ "winter",
                month == 3 | month == 4 | month == 5 ~ "spring",
                month == 6 | month == 7 | month == 8 ~ "summer",
                month == 9 | month == 10 | month == 11 ~ "fall"
        ))

util |> freq(season)

#8a: 29 of the bills are from fall.
#8b: 30 of the bills are from summer.

```

9) **BONUS (2 points):** Make a bar chart using the `geom_col()` to display the count or proportion of donors by `season`.\
\
First, make a new variable, `DonorStatus`. Set `DonorStatus` equal to 1 if the billee donated money to *Operation Fuel* and 0 otherwise. Check that you have created the `DonorStatus` variable correctly.\
\
Then, use `ggplot()` to create your graph. In your own words, describe the graph. Does donation vary by season?

```{r}
#| label: ans9
#| include: false
#| eval: false

util <- util |> 
        mutate(DonorStatus = case_when(
                donate == "yes" ~ 1,
                donate == "no" ~ 0
        ))

util |> freq(DonorStatus)

util |> 
        ggplot(aes(x = season, y = DonorStatus)) +
        geom_col()

#9: Most donations in the summer, least donations in the winter.

```

------------------------------------------------------------------------

# Submission

Submit your R script (named `LA-#_FirstName-LastName.R`) to Canvas.

Your R script should:

1)  Include commands and functions that are necessary to address all the questions in the assignment.
2)  Contain comments that answer the questions in the assignment.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Navigate back to Your Workspace on Posit Cloud.
-   Reopen your project.
-   Run the entire script line-by-line without editing it to ensure there are no errors.

::: callout-important
These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for all of your individual lab assignments.
:::
