---
title: 'LA-6: Examining and Managing Data (10 points)'
pagetitle: "COMM 3710: LA-6"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
urlcolor: blue
---

```{r setup, include=FALSE}
library(tidyverse)
library(summarytools)
hdata <- read.csv("HELP.csv")
knitr::opts_chunk$set(echo = FALSE)
```

In this assignment, you will put the skills you learned in LA-5 into practice.

Remember to ...

1) Set up your R script appropriately.
2) Include a comment for each line of code in your R script. This can be pseudocode or other comments.

---

# Step 1 {.tabset .tabset-fade}
## Instructions
Load the following packages in your current session of R:

- `tidyverse`
- `summarytools`
- `magrittr`
- `rstatix`

Read the [HELP](https://utah.instructure.com/files/117853074/download?download_frd=1) data set into R. Use the codebook to familiarize yourself with the variables in the data set.

## Hint
If you are having issues with `summarytools`, see the **Hint** in Step 2 of [LA-5](https://sarakyeo.github.io/COMM-3710/LA/LA-5.html).

## Codebook

```{r, codebook, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(tidyverse)
library(knitr)

codebook <- read_csv("https://raw.githubusercontent.com/sarakyeo/sarakyeo.github.io/master/COMM-3710/LA/HELP-codebook.csv")
codebook %>%
        kbl(., "html",
            caption = "Names and descriptions of variables in HELP dataset.") %>% 
        kable_styling(bootstrap_options = c("hover", "striped"),
                      full_width = FALSE) %>% 
        column_spec(2, width = "30em")

```


# Step 2 {.tabset .tabset-fade}
## Instructions
Patients with a mental component score (`mcs`) less than 20 are thought to be at extreme risk of returning to the detoxification unit within the next 12 months.

Make a new variable called `ExtremeMCS` and code it as `1` if a patient is at risk based on his/her mcs score and `0` otherwise. Then, answer the following questions.

a) How many patients are thought to be at high risk of returning to the detoxification unit within the next 12 months?
b) What percentage of patients are at low risk of returning to the detoxification unit within the next 12 months?

## Hint
Review LA-5 and how to use the `mutate()` and `case_when()` commands.


# Step 3
Make a new variable `SuicidalThought` based on the variable `g1b`. Have `1` indicate a patient has had suicidal thoughts and a `0` indicate they have not.

a) How many patients report having had suicidal thoughts?


# Step 4 
Make a new variable `HomelessStatus` based on the variable `homeless`. Have `1` indicate that a patient is homeless and `0` indicate that they are housed.

a) What percentage of patients are homeless?


# Step 5 {.tabset .tabset-fade}
## Instructions
Suppose we want to assess the overall risk of a patient and `ExtremeMCS`, `SuicidalThought`, and `HomelessStatus` are considered risk factors. Construct a new variable called `RiskTotal` that quantifies the number of risk factors for each patient. That is, make it a sum of these 3 variables.

a) What percentage of patients in the study have fewer than 3 risk factors?

## Hint
To create a new variable, `newVar`, that is the sum of three existing variables, `oldVar1`, `oldVar2`, and `oldVar3`, you can use the `mutate` command.
```{r step5, eval=FALSE, echo=TRUE}
dataframe <- dataframe %>% 
        rowwise %>% 
        mutate(newVar = oldVar1 + oldVar2 + oldVar3)
```
Then, you can check whether you have created the new variable, `newVar`, correctly by running a frequency distribution using the `freq` function.


# Submission
Submit your R script (`LA-6_FirstName-LastName.R`) to Canvas.

Your R script should:

1) Include code to install and load the packages.
2) Include code to set your working directory.
3) Contain comments and/or pseudocode for every single line of R code.
4) Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

- Save your script.
- Completely shut down RStudio or restart your R session.
- Reopen RStudio, open your `LA-6_FirstName-LastName.R` script.
- Run the entire script by clicking the "Run" button in the top right of the R script.

> These standards apply to all submissions in this course that require R scripts.
