---
title: "LA-6: Data Visualization (15 points)"
format:
   pdf:
      documentclass: article
      geometry:
         - top=20mm
         - left=20mm
         - right=20mm
         - heightrounded
      toc: false
      colorlinks: true
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(summarytools)
library(rstatix)
library(kableExtra)
library(formatR)

# Read data into R
util <- read_csv(here::here("COMM-3710", "LA","Utilities.csv"))

```

# Learning Outcomes

In this assignment, you will:

- Learn how to use `ggplot2`, which is a package included in the `tidyverse` suite of packages, to create visualizations using data.
- Practice using functions from LA-5, including `mutate()`, `case_when()`, `freq()`, and `descr()`.
- Combine conditions with `|` ("or").

::: callout-tip
Read all the instructions carefully before starting the assignment. Set up your R script as you did in LA-4 and LA-5 (name, uNID, date, assignment number).
:::

::: callout-important
**How to report your answers.** Write each answer as a comment in your R script, labeled with its question number (e.g., `# 3a: ...`). Round averages to two decimal places and include units (e.g., \$180.38).
:::

------------------------------------------------------------------------

# Instructions

1) Load the packages below. Then download the `Utilities.csv` dataset and its codebook from Canvas, upload the dataset to your Posit Cloud project (as you did in LA-4 and LA-5), and read it into R. Use the codebook to familiarize yourself with the data in this file.
   
   * `tidyverse`
   * `summarytools`

::: callout-note
**About this dataset.** Each row (case) is one monthly utility bill for **one household**, from 1999 to 2010. The bills include gas and electricity charges, how much gas and electricity the household used, and whether the household added a donation to *Operation Fuel* to that month's bill.
:::

2) Read and follow along with Chapter 5 of [this online book](https://sarakyeo.github.io/COMM3710-textbook/). Use the data provided in the chapter to follow the steps in Ch. 5.

:::callout-tip
Do not skip this step. It will help you understand `ggplot2`. You will use the same kinds of graphs from Ch. 5 below: a histogram (`geom_histogram()`), scatterplots (`geom_point()`), and a bar chart (`geom_col()`).
:::

3) Make a histogram to display the distribution of the household's total monthly bill (`totalbill`).

   a) Use the histogram you created and describe the distribution. Is the distribution skewed or symmetric? If skewed, is the distribution skewed positively or negatively?
   b) Where do most of the total monthly bills fall? Give a range in dollars (e.g., "between \$X and \$Y").

::: callout-tip
When you run `geom_histogram()`, R prints a message saying `` `stat_bin()` using `bins = 30` ``. This is not an error. If you want to control the width of the bars, you can use, for example, `geom_histogram(binwidth = 25)` to make each bar cover \$25.
:::
   
```{r}
#| label: ans3
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = totalbill)) +
        geom_histogram()

#3a: Distribution is skewed positively.
#3b: Most of the total monthly bills fall around $100 to $150.
# (48 of 117 bills are $100-$150; median = $134.50; middle half = $107-$193.
# Accept a reasonable range in this area.)

```

4) Make a scatterplot to display gas bill (`gasbill`) by month (`month`), with month on the x-axis. Read the tip below before you start, then use your graph to answer the following questions.

   a) Looking at your graph, which month appears to have the highest gas bills?
   b) Use R to calculate the average gas bill for the months of January and December. Which is higher?

::: callout-tip
`month` is stored as a number (1 = January, 12 = December), so by default the x-axis is labeled 2.5, 5, 7.5, 10. To label every month, add `scale_x_continuous(breaks = 1:12)` to your plot:

```
df |>
  ggplot(aes(x = month, y = variable)) +
  geom_point() +
  scale_x_continuous(breaks = 1:12)
```

For (b), use `filter()` and `descr()` as you did in LA-5, and read the `Mean` row.
:::
   
```{r}
#| label: ans4
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = month, y = gasbill)) +
        geom_point() +
        scale_x_continuous(breaks = 1:12)

#4a: January (accept December; the averages are $180.38 and $179.28)

util |> 
        filter(month == 1) |> 
        descr(gasbill) # M = 180.38

util |> 
        filter(month == 12) |> 
        descr(gasbill) # M = 179.28

#4b: Average gas bills are higher in January ($180.38) than December ($179.28).

```

5) Make a scatterplot to display electric bill (`elecbill`) by month (`month`), with month on the x-axis. Use your graph to answer the following questions.

   a) Looking at your graph, which month appears to have the highest electric bills?
   b) Use R to calculate the average electric bill for the months of September and December. Which is higher?
   
```{r}
#| label: ans5
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = month, y = elecbill)) +
        geom_point() +
        scale_x_continuous(breaks = 1:12)

#5a: September (highest average, $99.16; accept July, August, or September,
# the late-summer months: $90.68, $94.59, $99.16)

util |> 
        filter(month == 9) |> 
        descr(elecbill) # M = 99.16

util |> 
        filter(month == 12) |> 
        descr(elecbill) # M = 87.21

#5b: Average electricity bills are higher in September ($99.16) than December ($87.21).

```

6) Make a scatterplot to display the relationship between gas usage (`ccf`) and gas bill (`gasbill`). Place gas usage on the x-axis and monthly gas bill on the y-axis. Does there appear to be a relationship between gas usage and gas bill? If so, describe the relationship (e.g., does a relationship exist? Is it linear? Is there a positive or negative relationship between the variables?).

```{r}
#| label: ans6
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = ccf, y = gasbill)) +
        geom_point()

# Yes, there appears to be a relationship between gas bill and usage. There 
# appears to be a positive linear relationship between these two variables.

```

7) Make a scatterplot to show the relationship between electricity usage (`kwh`) and electric bill (`elecbill`). Place electricity usage on the x-axis and monthly electric bill on the y-axis. Does there appear to be a relationship between electricity usage and electric bill? If so, describe the relationship (e.g., does a relationship exist? Is it linear? Is there a positive or negative relationship between the variables?).

```{r}
#| label: ans7
#| include: false
#| eval: false

util |> 
        ggplot(aes(x = kwh, y = elecbill)) +
        geom_point()

# Yes, there appears to be a relationship between electricity bill and usage.
# There appears to be a positive linear relationship between these two variables.

```

8) Create a new categorical variable called `season`. Have the variable equal `winter` if the bill was from December, January, or February. Have the variable equal `spring` if the bill was from March, April, or May. Have the variable equal `summer` if the bill was from June, July, or August. Have the variable equal `fall` if the bill was from September, October, or November.\
\
Then, examine a frequency table of `season`. Read the note below before you start.

   a) How many bills in the data set are from fall?
   b) How many bills in the data set are from summer?

::: callout-note
Use `mutate()` and `case_when()` the same way you did in LA-5. This time, each season covers three months, so you need to combine conditions with `|`, which means **"or."** A row matches the condition if *any* part of it is `TRUE`. For example:

```
util <- util |>
  mutate(season = case_when(
    month == 12 | month == 1 | month == 2 ~ "winter",
    ...
  ))
```

Fill in the `...` with a line for each of the other three seasons. Remember to save the result back to `util` (start with `util <- util |> ...`), because you will use `season` again in #9.

(In LA-5, you used `&`, which means **"and"**: a row matches only if *every* part is `TRUE`. A bill can't be from December *and* January, so `&` won't work here.)
:::
   
```{r}
#| label: ans8
#| include: false
#| eval: false

util <- util |> 
        mutate(season = case_when(
                month == 12 | month == 1 | month == 2 ~ "winter",
                month == 3 | month == 4 | month == 5 ~ "spring",
                month == 6 | month == 7 | month == 8 ~ "summer",
                month == 9 | month == 10 | month == 11 ~ "fall"
        ))

util |> freq(season)

#8a: 29 of the bills are from fall.
#8b: 30 of the bills are from summer.

```

9) **BONUS (2 points):** Make a bar chart using `geom_col()` to display the number of bills that included a donation in each `season` (the variable you created in #8).\
\
First, make a new variable, `DonorStatus`. Set `DonorStatus` equal to 1 if the household donated money to *Operation Fuel* on that bill (`donate` is `"yes"`) and 0 otherwise. Check that you have created `DonorStatus` correctly by running `freq()` on both `donate` and `DonorStatus`: the number of `yes` bills should match the number of `1`s, and the number of `no` bills should match the number of `0`s.\
\
Then, use `ggplot()` to create your graph, with `season` on the x-axis and `DonorStatus` on the y-axis. In your own words, describe the graph. Does donation vary by season?

::: callout-tip
`geom_col()` adds up the y values for each category on the x-axis. Because `DonorStatus` is 1 for every bill with a donation and 0 otherwise, adding up `DonorStatus` within a season gives the **number of bills with a donation** in that season.
:::

```{r}
#| label: ans9
#| include: false
#| eval: false

util <- util |> 
        mutate(DonorStatus = case_when(
                donate == "yes" ~ 1,
                donate == "no" ~ 0
        ))

util |> freq(DonorStatus)

util |> 
        ggplot(aes(x = season, y = DonorStatus)) +
        geom_col()

#9: Most donations in the summer, least donations in the winter.
# (summer 11, fall 10, spring 9, winter 3; 33 yes / 84 no overall)

```

------------------------------------------------------------------------

# Submission

Submit your R script (named `LA-#_FirstName-LastName.R`) to Canvas.

Your R script should:

1)  Include commands and functions that are necessary to address all the questions in the assignment.
2)  Contain comments that answer the questions in the assignment.
3)  Run in its entirety without errors.

To ensure that your R script runs without errors, you should:

-   Save your script.
-   Navigate back to Your Workspace on Posit Cloud.
-   Reopen your project.
-   Run the entire script line-by-line without editing it to ensure there are no errors.

::: callout-important
These standards apply to all submissions in this course that require R scripts. You should follow these instructions for preparation, naming, and saving of your R script for all of your individual lab assignments.
:::
