---
title: 'LA.4: Working with Data (10 points)'
pagetitle: "COMM 3710: LA.4"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
urlcolor: blue
---

```{r setup, include=FALSE}
library(tidyverse)
library(descr)
hdata <- read.csv("HELP.csv")
knitr::opts_chunk$set(echo = FALSE)
```

# Resources

- [Video 05: Data Management](https://youtu.be/b9Bh3K5bi0I)
- [COMM 3710: Getting Started with R](https://sarakyeo.github.io/COMM-3710-bookdown/docs/index.html)

---

# Data

For this assignment, you will be working with the [Health Evaluation and Linkage to Primary Care (HELP) data set](https://utah.instructure.com/files/108105219/download?download_frd=1). Submit your answers as a PDF document on Canvas.

The HELP study was a clinical trial for adult inpatients recruited from a detoxification unit. Patients with no primary care physician were randomized to receive a multidisciplinary assessment and a brief motivational intervention or usual care, with the goal of linking them to primary medical care.

The data set contains 453 observations on the following variables:

* `age`: Subject age at baseline (in years)
* `anysub`: Use of any substance post-detox (a factor with levels no yes)
* `cesd`: Center for Epidemiologic Studies Depression measure at baseline (high scores indicate more depressive symptoms)
* `d1`: lifetime number of hospitalizations for medical problems (measured at baseline)
* `daysanysub`: time (in days) to first use of any substance post-detox
* `dayslink`: time (in days) to linkage to primary care
* `drugrisk`: Risk Assessment Battery drug risk scale at baseline
* `e2b`: number of times in past 6 months entered a detox program (measured at baseline)
* `female`: 0 for male, 1 for female
* `sex`: a factor with levels male female
* `g1b`: experienced serious thoughts of suicide in last 30 days (measured at baseline; a factor with levels no yes)
* `homeless`: housing status: a factor with levels housed homeless
* `i1`: average number of drinks (standard units) consumed per day, in the past 30 days (measured at baseline)
* `i2`: maximum number of drinks (standard units) consumed per day, in the past 30 days (measured at baseline)
* `id`: subject identifier
* `indtot`: Inventory of Drug Use Consequences (InDUC) total score (measured at baseline)
* `linkstatus`: post-detox linkage to primary care (0 = no, 1 = yes)
* `link`: post-detox linkage to primary care: no yes
* `mcs`: SF-36 Mental Component Score (measured at baseline, lower scores indicate worse status)
* `pcs`: SF-36 Physical Component Score (measured at baseline, lower scores indicate worse status)
* `pss_fr`: perceived social support by friends (measured at baseline, higher scores indicate more support)
* `racegrp`: race/ethnicity: levels black hispanic other white
* `satreat`: any BSAS substance abuse treatment at baseline (no yes)
* `sexrisk`: Risk Assessment Battery sex risk score (measured at baseline)
* `substance`: primary substance of abuse (alcohol cocaine heroin)
* `treat`: randomized to HELP clinic (no yes)

---

# Instructions

## Step 1 {.tabset .tabset-fade .tabset-pills}
### Question
Load in the [HELP data set](https://utah.instructure.com/files/108105219/download?download_frd=1) that we used in [LA.3](https://sarakyeo.github.io/COMM-3710/LA/LA.3.html).


## Step 2 {.tabset .tabset-fade .tabset-pills}
### Question
Patients with a mental component score (`mcs`) less than 20 are thought to be at extreme risk of returning to the detoxification unit within the next 12 months.

Make a new variable called `ExtremeMCS` and code it as `1` if a patient is at risk based on his/her mcs score and `0` otherwise.

### Hint
In addition to the method shown in the video, you can create this new variable using the `ifelse` command.
   
Figure out what the arguments (the text and numbers within the parentheses) in the `ifelse` function mean by typing `?ifelse` into the Console in RStudio and reading the documentation.

The code below creates a new variable called `alcohol` based on the existing variable, `substance`, in the HELP data set, which has been read into a data frame called `hdata`. It recodes the `substance` variable into a new variable that assigns `1` to any patient whose primary substance of abuse is alcohol and `0` to patients whose primary substance of abuse is cocaine or heroine. 

```{r recode, eval=TRUE, echo=TRUE, prompt=TRUE}
hdata$alcohol <- ifelse(hdata$substance == "alcohol", 1, 0)
freq(hdata$alcohol)
```


## Step 3 {.tabset .tabset-fade .tabset-pills}
Make a new variable `SuicidalThought` based on the variable `g1b`. Have `1` indicate a patient has had suicidal thoughts and a `1` indicate they have not.


## Step 4 {.tabset .tabset-fade .tabset-pills}
Make a new variable `HomelessStatus` based on the variable `homeless`. Have `1` indicate that a patient is homeless and `0` indicate that they are housed.


## Step 5 {.tabset .tabset-fade .tabset-pills}
### Question
Suppose we want to assess the overall risk a patient has to return to the detoxification unit and `ExtremeMCS`, `SuicidalThought`, and `HomelessStatus` are considered risk factors. Construct a new variable called `RiskTotal` which computes the number of risk factors a particular patient has. That is, make it a sum of these 3 variables.

### Hint
You want to create a new variable called `RiskTotal`, which is the sum of `ExtremeMCS`, `SuicidalThought`, and `HomelessStatus`. First, try the most straightforward way, which is to use this general formula:

```
Dataframe$NewVariable <- Dataframe$Var1 + Dataframe$Var2 + Dataframe$Var3 
```

Like most things in R, there are many other ways to do this. One way is to transform the data using the command, `transform`. Get help with `transform` by typing `?transform` into the Console and reading the documentation.


## Step 6 {.tabset .tabset-fade .tabset-pills}
### Question
Make frequency tables of `ExtremeMCS`, `SuicidalThought`, and `RiskTotal`. Answer the following questions.

a. How many patients are thought to be at risk based on their `mcs` score?
b. What percentage of patients have experienced suicidal thoughts?
c. What percentage of patients in the study have fewer than 3 risk factors? 

### Hint
Refer to LA.3 to review creating frequency tables.


## Submission {.tabset .tabset-fade .tabset-pills}
Submit a **.R file** containing the following on Canvas:

- R code/commands used to complete this assignment. Your R script should include annotations that explain your thought process as you work through the assignment.
- Clearly labeled answers to questions as comments/annotations in the R script.
