---
title: "LA.7: Practice Data Analysis (10 points)"
output:
  pdf_document: default
  html_document: default
  word_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
# Install packages
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("descr")
# install.packages("ggpubr")
# install.packages("rstatix")
# install.packages("rmarkdown")
# install.packages("formatR")

# Load packages
library(tidyr)
library(dplyr)
library(ggplot2)
library(descr)
library(ggpubr)
library(rstatix)
library(rmarkdown)
library(formatR)
```

## Overview
Before starting this assignment, review the following videos, if necessary:

1. Hypothesis testing: https://youtu.be/08uylSp-CuI
2. Analysis of variance (ANOVA) in R: https://youtu.be/2uQCUBtqCfM
3. Chi-square test of independence in R: https://youtu.be/Dnlp9vswwEk
4. Correlation: https://youtu.be/lf4HfHx2ASs

Remember that help with R is also available through Slack (see Canvas announcements) and on [this website](https://rpubs.com/yesun/Comm3710_helpwithR).

[Data for this assignment comes from the 1985 Current Population Survey (CPS)](https://utah.instructure.com/files/102627573/download?download_frd=1). The CPS is used to supplement census information between census years. These data consist of a random sample of persons from the survey, with information on wages and other characteristics of the workers, including sex, number of years of education, years of work experience, occupational status, region of residence and union membership.

A data frame with 534 observations on the following variables:

* `wage`: wage (US dollars per hour)
*	`educ`: number of years of education
*	`race`: a factor with levels NW (nonwhite) or W (white)
*	`sex`: a factor with levels FM
*	`hispanic`: a factor with levels Hisp or NH
*	`south`: a factor with levels NSS
*	`married`: a factor with levels Married or Single
*	`exper`: number of years of work experience (inferred from age and educ)
*	`union`: a factor with levels NotUnion or Union
*	`age`: age in years
*	`sector`: a factor with levels clerical const manag manuf other prof sales service
*	`satisfaction`: 1 implies participant satisfied with current employment, 0 otherwise

## Instructions
1. Familiarize yourself with the CPS documentation above. Then, import/load the [CPS data set](https://utah.instructure.com/files/102627573/download?download_frd=1).

    ```{r, include=FALSE, tidy=TRUE}
    # 1
    cps <- read.csv("CPS.csv")
```

2. Find the mean wage earned per hour for males and females.
      a. What is the mean wage earned per hour for males and females? <!--M = $9.99, F = $7.88-->
      
    ```{r, include=FALSE, tidy = TRUE}
    # 2
    mwage <- mean(cps$wage[cps$sex == "M"])
    fwage <- mean(cps$wage[cps$sex == "F"])
```

3. Implement the appropriate statistical test to determine whether males have significantly higher wages earned per hour than females. Use the chart from LA.6 to guide your choice of statistical test.
      a. What statistical test did you use? <!--Independent samples t-test-->
      b. What is the value of the test statistic? <!--t = -4.84, p < .001-->
      c. What conclusions can be drawn? <!--There is a statistically significant difference in mean wage per hour between men and women-->
      
    ```{r, include = FALSE, tidy = TRUE}
    #3
    t.test(cps$wage ~ cps$sex, var.equal = TRUE)
```

4. Implement the appropriate statistical test to determine whether there is a significant linear relationship between wages earned per hour and number of years of work experience.
      a. What statistical test did you use? <!--Pearson's correlation-->
      b. What is the value of the test statistic? <!--Pearson's r = .09, p = .044-->
      c. What conclusions can be drawn? <!-- There is a statistically significant relationship between the variables that is weak and positive-->
      
    ```{r, include = FALSE, tidy = TRUE}
    # 4
    res.cor <- cor.test(cps$wage, cps$exper, method = "pearson")
    res.cor
```

5. Implement the appropriate statistical test to determine whether job satisfaction varies significantly among the different job sectors.
      a. What statistical test did you use? <!--Chi-squared test-->
      b. What is the value of the test statistic? <!--X-squared = 3.65, df = 7, p = .82-->
      c. What conclusions can be drawn? <!--Job satisfaction and job sector are not related-->
      
    ```{r, include = FALSE, tidy = TRUE}
    # 5
    tbl <- table(cps$sector, cps$satisfaction)
    chisq.test(tbl)
```

6. Submit your answers as a PDF on Canvas.
