---
title: "LA.8: Practice Data Analysis (10 points)"
output:
  pdf_document: default
  html_document: default
  word_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
# Install packages
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("descr")
# install.packages("ggpubr")
# install.packages("rstatix")
# install.packages("rmarkdown")
# install.packages("formatR")

# Load packages
library(tidyr)
library(dplyr)
library(ggplot2)
library(descr)
library(ggpubr)
library(rstatix)
library(rmarkdown)
library(formatR)
```

## Overview
Video resources for data analysis with R:

1. Hypothesis testing: https://youtu.be/08uylSp-CuI
2. Analysis of variance (ANOVA) in R: https://youtu.be/2uQCUBtqCfM
3. Chi-square test of independence in R: https://youtu.be/Dnlp9vswwEk
4. Correlation: https://youtu.be/lf4HfHx2ASs

You will use the [Ithaca.csv](https://utah.instructure.com/files/102628202/download?download_frd=1) data set for this assignment. These data were collected by students in a research methods course at Cornell University for teaching purposes.

A data frame with 465 observations on the following variables (Note: 99 = “Don’t know” unless otherwise specified):

*	`tvinat`: Attention to international news (0 = “Little attention,” 10 = “Very close attention”)
*	`tvnaat`: Attention to national news (0 = “Little attention,” 10 = “Very close attention”)
*	`tvhardat`: Attention to news about politics and economy (0 = “Little attention,” 10 = “Very close attention”)
*	`tvsoftat`: Attention to soft news (0 = “Little attention,” 10 = “Very close attention”)
*	`tvcrat`: Attention to crime dramas (0 = “Little attention,” 10 = “Very close attention”)
*	`tvcomat`: Attention to ctvomedies (0 = “Little attention,” 10 = “Very close attention”)
*	`educ`: Highest year of school completed
*	`age`: Age of respondent
*	`marit`: Marital status (1 = “single,” 2 = “married,” 3 = “divorced/separated,” 4 = “widowed,” 5 = “other,” 9 = “Don’t know/NA”)
*	`sex`: 1 =  0 = “female,” 1 = “male”

## Instructions
1. Familiarize yourself with the [Ithaca.csv](https://utah.instructure.com/files/102628202/download?download_frd=1) codebook above. Then, import/load the data set.
    ```{r, include = FALSE, tidy = TRUE}
# 1
ithaca <- read.csv("Ithaca.csv")
```

2. Conduct data management as necessary on these variables: `tvnaat`, `tvhardat`, `tvsoftat`, `sex`, `marit`. E.g., you might have to remove missing data or recode "Don't know" responses as NA.
    ```{r, include = FALSE, tidy = TRUE}
# 2: Managing ordinal variables: Recoding "Don't know" responses as missing data, i.e., "NA."
freq(ithaca$tvnaat)
ithaca$ctvnaat <- ithaca$tvnaat
ithaca$ctvnaat[ithaca$tvnaat == 99] <- NA
freq(ithaca$ctvnaat)

freq(ithaca$tvhardat)
ithaca$ctvhardat <- ithaca$tvhardat
ithaca$ctvhardat[ithaca$tvhardat == 99] <- NA
freq(ithaca$ctvhardat)

freq(ithaca$tvsoftat)
ithaca$ctvsoftat <- ithaca$tvsoftat
ithaca$ctvsoftat[ithaca$tvsoftat == 99] <- NA
freq(ithaca$ctvsoftat)

# Recoding nominal variables as factors
freq(ithaca$sex)
ithaca$csex <- ithaca$sex
ithaca$csex[ithaca$sex == 9] <- NA
ithaca$csex[ithaca$sex == 1] <- "M"
ithaca$csex[ithaca$sex == 0] <- "F"
freq(ithaca$csex)

freq(ithaca$marit) # In this case, the values 5 and 9 are not meaningful in this variable
ithaca$cmarit <- ithaca$marit
ithaca$cmarit[ithaca$marit == 1] <- "single"
ithaca$cmarit[ithaca$marit == 2] <- "married"
ithaca$cmarit[ithaca$marit == 3] <- "divorced/separated"
ithaca$cmarit[ithaca$marit == 4] <- "widowed"
ithaca$cmarit[ithaca$marit == 5] <- NA
ithaca$cmarit[ithaca$marit == 9] <- NA
freq(ithaca$cmarit)
```

3. Find the mean attention to national TV news for males and females.
      a. What is the mean attention to national TV news for men and women? <!-- `men: M = 5.181, women: M = 5.176` -->
    ```{r, include = FALSE, tidy = TRUE}
# 3
FMean <- mean(ithaca$ctvnaat[ithaca$csex == "F"], na.rm = TRUE)
MMean <- mean(ithaca$ctvnaat[ithaca$csex == "M"], na.rm = TRUE)
FMean
MMean
```

4. Implement the appropriate statistical test to determine whether men spend significantly more time watching national TV news compared to women.
      a. What statistical test did you use? <!-- `Independent samples t-test` -->
      b. What is the value of the test statistic and the *p*-value? <!-- `t = .016, p = .988` -->
      c. What conclusions can be drawn? <!-- `Attention to national news on TV is not significantly different between men and women.` -->
    ```{r, include = FALSE, tidy = TRUE}
# 4
res.t <- t.test(ithaca$ctvnaat~ithaca$csex, var.equal = TRUE)
res.t
```

5. Implement the appropriate statistical test to determine whether there is a significant linear relationship between attention to national TV news and attention to news about politics and economy.
      a. What statistical test did you use? <!-- `Pearson's correlation` -->
      b. What is the value of the test statistic and the *p*-value? <!-- `Pearson's r = .649, p < .001` -->
      c. What conclusions can be drawn? <!-- `There is a statistically significant positive relationship between the variables.` -->
    ```{r, include = FALSE, tidy = TRUE}
# 5
res.cor <- cor.test(ithaca$ctvnaat, ithaca$ctvhardat, method = "pearson")
res.cor
```

6. Implement the appropriate statistical test to determine whether attention to soft news varies significantly depending on marital status.
      a. What statistical test did you use? <!--`One-way analysis of variance (ANOVA)`-->
      b. What is the value of the test statistic and the *p*-value? <!-- `F = 5.197, p = .002` -->
      c. What conclusions can be drawn? <!-- `There is a significant different in attention to soft news by marital status.`-->
      <!-- d. Are *post hoc* tests necessary? If so, what are the results? Yes. Only significant difference in attention to soft news between single and divorced/separated R; pwc <- ithaca %>% tukey_hsd(ctvsoftat ~ cmarit); pwc-->
```{r warning=FALSE, include=FALSE, tidy = TRUE}
# 6
res.aov <- ithaca %>% anova_test(ctvsoftat ~ cmarit)
res.aov
```
